---
title: "10 - Tibble"
author: "Andrea Pedot"
date: "24/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Exercises

1.  How can you tell if an object is a tibble?
    (Hint: try printing `mtcars`, which is a regular data frame).
    
```{r}
is_tibble(mtcars)
```

2.  Compare and contrast the following operations on a `data.frame` and equivalent tibble.
    What is different?
    Why might the default data frame behaviours cause you frustration?

    ```{r, eval = FALSE}
    df <- data.frame(abc = 1, xyz = "a")
    df$x # this partial matching is a problem
    df[, "xyz"] #it is long to write, why does it return a list?
    df[, c("abc", "xyz")] # why does it return a data.frame
    
    #try with tibble
    df <- as_tibble(data.frame(abc = 1, xyz = "a"))
    df$x # ERROR (unknown column)
    df[, "xyz"] # return 1*1 tibble
    df[, c("abc", "xyz")] # return 1*2 tibble
    ```

3.  If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, how can you extract the reference variable from a tibble?

```{r}
var <- "mpg"

mtcars$var # does not work in data.frame
mtcars[, var] # output is a list

tibblecars <- as_tibble(mtcars)
tibblecars$var # does not work in data.frame
tibblecars[, var] # output is a tibble
```

4.  Practice referring to non-syntactic names in the following data frame by:

    ```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
    ```

    1.  Extracting the variable called `1`.

    ```{r}
annoying$`1`
```

    2.  Plotting a scatterplot of `1` vs `2`.
    
```{r}
ggplot(annoying,
       aes( x = `1`,
            y = `2` )) +
  geom_point()
```

    3.  Creating a new column called `3` which is `2` divided by `1`.
    
```{r}
(
annoying <- annoying %>% 
  mutate( `3` = `2` /  `1`)
)
```

    4.  Renaming the columns to `one`, `two` and `three`.

```{r}
annoying %>% 
  rename( `one` = `1`,
          `two`  = `2` ,
          `three`  = `3` )
```

    
5.  What does `tibble::enframe()` do?
    When might you use it?
    
Converts named atomic vectors or lists to one- or two-column.

enframe(x, name = "name", value = "value") with x as an atomic vector, name as the name column (if NULL it will produce a one-column tibble) and value for values.

It will be used to quickly convert vectors in tibble for analytical purposes.

6.  What option controls how many additional column names are printed at the footer of a tibble?

```{r, eval = FALSE}
tibble.max_extra_cols # Number of extra columns printed in reduced form. Default: 100
```

### Exercises

1.  What function would you use to read a file where fields were separated with\
    "\|"?
    
```{r, eval = FALSE}
read_delim( file, delim = "|")
```
    
2.  Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` and `read_tsv()` have in common?

-  col_names = TRUE as the first row of the input will be used as the column names, otherwise column names will be generated automatically: X1, X2, X3. If a character vector, values will be used as the names of the columns
-  col_types = NULL, all column types will be imputed from the first 1000 rows on the input
-  locale = default_locale(), it controls things like the default time zone, encoding, decimal mark, big mark
-  na = c("", "NA") is the character vector of strings to interpret as missing values
-  quoted_na = TRUE, if missing values inside quotes should be treated as missing values
-  quote = "\"" is the single character used to quote strings
-  trim_ws = TRUE, as leading and trailing whitespace should be trimmed from each field before parsing it
-  n_max = Inf, Maximum number of records to read
-  guess_max = min(1000, n_max), Maximum number of records to use for guessing column types
-  progress = show_progress() to display a progress bar
-  skip_empty_rows = TRUE if blank rows should be ignored altogether

3.  What are the most important arguments to `read_fwf()`?

- file,
- col_positions, as created by fwf_empty(), fwf_widths() or fwf_positions()

You can specify column positions in several ways, given that fwf_sample contains the fixed width file:

- Guess based on position of empty columns read_fwf(fwf_sample, fwf_empty(fwf_sample, col_names = c("first", "last", "state", "ssn")))
- A vector of field widths, read_fwf(fwf_sample, fwf_widths(c(20, 10, 12), c("name", "state", "ssn")))
- Paired vectors of start and end positions, read_fwf(fwf_sample, fwf_positions(c(1, 30), c(20, 42), c("name", "ssn")))
- Named arguments with start and end positions, read_fwf(fwf_sample, fwf_cols(name = c(1, 20), ssn = c(30, 42)))
- Named arguments with column widths, read_fwf(fwf_sample, fwf_cols(name = 20, state = 10, ssn = 12))

4.  Sometimes strings in a CSV file contain commas.
    To prevent them from causing problems they need to be surrounded by a quoting character, like `"` or `'`. By default, `read_csv()` assumes that the quoting character will be `"`.
    What argument to `read_csv()` do you need to specify to read the following text into a data frame?

    ```{r, eval = FALSE}
( x <- read_csv("x,y\n1,'a,b'", quote = "'") )
    ```

5.  Identify what is wrong with each of the following inline CSV files.
    What happens when you run the code?

    ```{r, eval = FALSE}
    read_csv("a,b\n1,2,3\n4,5,6") # 2 columns but 3 columns literal data
    read_csv("a,b,c\n1,2\n1,2,3,4") # 3 columns but 2 or 4 columns literal data
    read_csv("a,b\n\"1") # closing quote at end of file + 2 columns but 1 columns literal data
    read_csv("a,b\n1,2\na,b") # int parsed as char
    read_csv("a;b\n1;3") # different delimiter
    ```
