---
title: "12 - Tidy data"
author: "Andrea Pedot"
date: "12/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Exercises

1.  Using prose, describe how the variables and observations are organised in each of the sample tables.

- Table 1 each row represent a country-year observation, with four variables in four columns

- Table 2 shows in each row either population or cases, according to the "Type" column which does not represent a variable, while count contains that information

- Table 3 merges in one column both the information on cases and total population as a rate. The third column is character vector, so not easily readable for a script

- Table 4a and 4b represent one row for each observation and one column for each variable, but they are more difficult to combine

2.  Compute the `rate` for `table2`, and `table4a` + `table4b`.
    You will need to perform four operations:

    a.  Extract the number of TB cases per country per year.
    b.  Extract the matching population per country per year.
    c.  Divide cases by population, and multiply by 10000.
    d.  Store back in the appropriate place.

```{r}
# from table2

cases <- table2 %>% 
  group_by(country, year) %>% 
  filter( type == "cases") %>% 
  select(count)

population <- table2 %>% 
  group_by(country, year) %>% 
  filter( type == "population") %>% 
  select(count)

rate <- tibble(
  country = table4c$country,
  year = table4c$year,
  type = "rate",
  count = cases$count / population$count * 1e4
)

rbind(table2, rate) %>% 
  arrange(country, year)

# from table4
table4c <- tibble(
  country = table4a$country,
  `1999` =  table4a$`1999` / table4b$`1999`  * 1e4,
  `2000` =  table4a$`2000` / table4b$`2000`  * 1e4,
)
```


    Which representation is easiest to work with? Table1
    
    Which is hardest? Table2
    
    Why? Data can not be easily formatted

3.  Recreate the plot showing change in cases over time using `table2` instead of `table1`.
    What do you need to do first? Filter to specific rows of the dataframe

```{r}
ggplot(filter(table2, type == "cases"), aes(year, count)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```


## Exercises
1.  Why are `pivot_longer()` and `pivot_wider()` not perfectly symmetrical?\
    Carefully consider the following example:

```{r, eval = FALSE}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half   = c(   1,    2,    1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>%
  pivot_wider(names_from = year, values_from = return) %>%
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

"Year" variable now is a character vector.

    (Hint: look at the variable types and think about column *names*.)

    `pivot_longer()` has a `names_ptypes` argument, e.g. `names_ptypes = list(year = double())`.
    What does it do?
    
A list of column name-prototype pairs. A prototype (or ptype for short) is a zero-length vector (like integer() or numeric()) that defines the type, class, and attributes of a vector. Use these arguments to confirm that the created columns are the types that you expect.

If not specified, the type of the columns generated from names_to will be character, and the type of the variables generated from values_to will be the common type of the input columns used to generate them.

2.  Why does this code fail?

```{r, error = TRUE}
table4a %>%
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
```

It should be with id_cols = c(`1999`, `2000`)

3.  What would happen if you widen this table?
    Why?
    How could you add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|-------
  "Phillip Woods",   "age",        45,
  "Phillip Woods",   "height",    186,
  "Phillip Woods",   "age",        50,
  "Jessica Cordero", "age",        37,
  "Jessica Cordero", "height",    156
)

people %>% 
  group_by( name, names) %>% 
  mutate( i = row_number()) %>% 
  pivot_wider( names_from = names,
               values_from = values,
               names_repair = "minimal")
```
The output is a vector for each of the values, since row 1 and 3 have the same identifiers.

4.  The simple tibble below summarizes information on whether employees at a small company know how to drive and whether they prefer a position where they will need to drive daily for sales calls.
    Tidy the table to get it into a format where each observation is an employee.
    Do you need to make it wider or longer?
    What are the variables?

```{r}
employees <- tribble(
  ~know_drive, ~prefer, ~not_prefer,
  "yes",       20,      10,
  "no",        NA,      12
)

employees %>% 
  pivot_longer(c("prefer", "not_prefer"),
               names_to = "preference",
               values_to = "count",
               values_drop_na = TRUE)
```

5.  One way of summarising the distribution of one categorical variable based on the levels of another is using `dplyr::count()`, e.g. the following gives the distribution of `drv` (type of drive train) for each level of `cyl` (number of cylinders) for cars in the `mpg` dataset.

    ```{r}
    mpg %>%
      count(cyl, drv)
    ```

    A contingency table is another way commonly used way of summarising this information.
    Use one of the pivoting functions to construct the contingency table shown below based on the output above.

    ```{r echo = FALSE}
    mpg %>%
      count(cyl, drv) %>%
      pivot_wider(names_from = drv, values_from = n)
    ```


