---
title: "5 - Transformation"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and dataframe

```{r, message = FALSE}
library(tidyverse)
library(nycflights13)
library(zoo) # for computing running average
#library(fitdistrplus) #for beta distribution
```

### Exercises

1.  Find all flights that

1. Had an arrival delay of two or more hours
```{r}
flights %>% filter(arr_delay >= 120)
```
    
2. Flew to Houston (`IAH` or `HOU`)
```{r}
flights %>% filter(dest %in% c("IAH", "HOU"))
```
    
3. Were operated by United, American, or Delta
  First we have to check which code has each airline
```{r}
airlines
```

  Then we can filter for them
```{r}
flights %>% filter(carrier %in% c("DL", "AA", "UA"))
```
    
4. Departed in summer (July, August, and September)
```{r}
flights %>% filter(month %in% 7:9)
```
    
5. Arrived more than two hours late, but didn't leave late
```{r}
flights %>% filter(arr_delay >= 120 & dep_delay <= 0)
```
    
6. Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
flights %>% filter(dep_delay >= 60 & arr_delay < dep_delay - 30)
```

7. Departed between midnight and 6am (inclusive)
```{r}
flights %>% filter(dep_time %in% 0000:0600)
```
    



2.  Another useful dplyr filtering helper is `between()`. What does it do?
    Can you use it to simplify the code needed to answer the previous 
    challenges?
```{r}
flights %>% filter(between(month, 7, 9))
```

```{r}
flights %>% filter(between(dep_time, 0000, 0600))
```
    

3.  How many flights have a missing `dep_time`? What other variables are 
    missing? What might these rows represent?

```{r}
flights %>% filter(is.na(dep_time))
```

    8255 flights are missing dep_time.
    Maybe they are cancelled flights, we can check if (missing dep_time - missing air_time) is empty to confirm this hypothesis.
    
```{r}
flights %>% filter(is.na(dep_time) & !is.na(air_time))
```

4.  Why is `NA ^ 0` not missing?Why is `NA | TRUE` not missing?
    Why is `FALSE & NA` not missing? Can you figure out the general
    rule?  (`NA * 0` is a tricky counterexample!)
    
The NA could be any number. If the number is finite, then the result of the multiplication will be 0, the exponential zero will be 1 and logical operation will show that result under any circumstance.
    
Counterexample: However, if the the number is Inf, then the result of the multiplication will be NaN. We don't know if the multiplication results in 0 or NaN so the result is given as NA.
    
    

### Exercises

1.  How could you use `arrange()` to sort all missing values to the start?
    (Hint: use `is.na()`).
    
  First we have to check where the NAs are
```{r}
sapply(flights, function(x) sum(is.na(x)) )
```
  Then for each column with NA we sort the dataset (the more NAs, the higher is the position)
```{r}
flights %>% arrange(desc(is.na(dep_time) +
                         is.na(arr_time) +
                         is.na(tailnum)  +
                         is.na(air_time) ))
```
    
2.  Sort `flights` to find the most delayed flights. Find the flights that
    left earliest.
    
    [Note that earliest refers here to the absolute time and it is not compared to the sched_dep_time]
    
```{r}
arrange(flights, desc(dep_delay), dep_time)
```
    

3.  Sort `flights` to find the fastest (highest speed) flights.
```{r}
arrange(flights, desc( distance/air_time ))
```

4.  Which flights travelled the farthest? Which travelled the shortest?

Farthest flight
```{r}
arrange(flights, desc(distance)) %>% head(1)
```

Shortest non cancelled flight (second row)
```{r}
arrange(flights, distance) %>% head(2)
```



### Exercises

1.  Brainstorm as many ways as possible to select `dep_time`, `dep_delay`,
    `arr_time`, and `arr_delay` from `flights`.
```{r}
flights %>% select(4, 6, 7, 9)
flights %>% select(starts_with("dep_"), starts_with("arr_"))
flights %>% select(starts_with("dep_")| starts_with("arr_"))
flights %>% select( (contains("time") | contains("delay") ) &
                    (contains("dep")  | contains("arr")   ) & 
                    !contains("sched")                      )
```
    
2.  What happens if you include the name of a variable multiple times in
    a `select()` call?
It doesn't get replicated, as in
```{r}
flights %>% select(arr_time, arr_time, arr_time)
```
  
3.  What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?
    
    It selects all the variables apart from the ones in any_of
    ```{r}
    vars <- c("year", "month", "day", "dep_delay", "arr_delay")
    flights %>% select(any_of(vars))
    ```
    
4.  Does the result of running the following code surprise you?  How do the
    select helpers deal with case by default? How can you change that default?

'r contains(match, ignore.case = TRUE, vars = NULL)'

    ```{r, eval = FALSE}
    select(flights, contains("TIME"))
    ```
    
```{r}
    select(flights, contains("TIME", ignore.case = FALSE))
```
    


### Exercises

1.  Currently `dep_time` and `sched_dep_time` are convenient to look at, but
    hard to compute with because they're not really continuous numbers. 
    Convert them to a more convenient representation of number of minutes
    since midnight.

```{r}
flights <- flights %>%
  mutate(
    dep_time_min = 60* dep_time %/% 100 + dep_time %% 100,
    arr_time_min = 60* arr_time %/% 100 + arr_time %% 100,
    sched_dep_time_min = 60* sched_dep_time %/% 100 + sched_dep_time %% 100
    )

```
    
1.  Compare `air_time` with `arr_time - dep_time`. What do you expect to see?
    What do you see? What do you need to do to fix it?

```{r}
flights <- flights %>%
  mutate( flight_difference = air_time - (arr_time - dep_time_min))

ggplot(flights) +
  geom_density(aes(flights$flight_difference))
# wrong
```

```{r}

flights <- flights %>%
  mutate( flight_difference_min = air_time - (arr_time_min - dep_time_min))

ggplot(flights) +
  geom_density(aes(flights$flight_difference_min)) +
  geom_vline(xintercept = c(60*(-2:3)), color = "red")

```

 There is the combined effect of time zones.
 
 Moreover, the time now is calculated using minutes since midnight to avoid effect of uncorrectly substract HHMM integers.
    
1.  Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you
    expect those three numbers to be related?

```{r}
flights <- flights %>%
  mutate( dep_difference = dep_delay - (dep_time_min - sched_dep_time_min))

ggplot(flights) +
  geom_density(aes(flights$flight_difference)) +
  geom_vline(xintercept = median(flights$flight_difference, na.rm = TRUE),
             color = "red")
```

They should be related as `dep_delay = (dep_time - sched_dep_time)`


1.  Find the 10 most delayed flights using a ranking function. How do you want 
    to handle ties? Carefully read the documentation for `min_rank()`.

```{r}
flights %>%
  select(flight, dep_time_min, sched_dep_time_min) %>%
  mutate(dep_delay_rank = min_rank(desc(dep_time_min - sched_dep_time_min))) %>%
  arrange(dep_delay_rank) %>%
  head(10)

```


1.  What does `1:3 + 1:10` return? Why?

Error because longer object length is not a multiple of shorter object length

1.  What trigonometric functions does R provide?

cos(x), sin(x), tan(x), acos(x), asin(x), atan(x), atan2(y,x)



### Exercises

1.  Brainstorm at least 5 different ways to assess the typical delay 
    characteristics of a group of flights. Consider the following scenarios:
    
    * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of 
      the time.
      
    * A flight is always 10 minutes late.

    * A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of 
      the time.
      
    * 99% of the time a flight is on time. 1% of the time it's 2 hours late.
    
    Which is more important: arrival delay or departure delay?
    
It is more important the arrival delay.

Basic assessment involves median and average delay as well as variance or IQR (filtering arr_delay > 0), count and proportion of flights with a no delay, small_delay (<20') and significant delay (>20').

geom_dist can provide useful insights on delay pattern for that particolar group.

```{r}

flights %>%
  group_by( carrier ) %>%
  summarize( prop_delay = mean( arr_delay > 0, na.rm = TRUE)) 
```

```{r}
flights %>%
  group_by( carrier ) %>%
  filter ( arr_delay > 0 ) %>% 
  group_by( carrier ) %>% 
  summarize( mean_delay = mean(arr_delay, na.rm = TRUE),
             sd_delay = sd( arr_delay, na.rm = TRUE),
             median_delay = median( arr_delay, na.rm = TRUE),
             iqr_delay = IQR( arr_delay, na.rm = TRUE)) %>% 
  arrange(desc(mean_delay))
```

```{r}
ggplot(filter(flights, arr_delay > 0)) +
  geom_density( aes( x = arr_delay ), color = "grey50", size = 2) +
  geom_density( aes( x = arr_delay, color = carrier )) +
  scale_x_log10()
```

1.  Come up with another approach that will give you the same output as 
    `not_cancelled %>% count(dest)` and 
    `not_cancelled %>% count(tailnum, wt = distance)` (without using 
    `count()`).
    
```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(dest) %>% 
  summarise( n() )
```

```{r}
not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise( sum(distance) )
```

1.  Our definition of cancelled flights (`is.na(dep_delay) | is.na(arr_delay)`
    ) is slightly suboptimal. Why? Which is the most important column?
    
  We are interested only in arrived flights, which accounts also for deviations or plane crashes

1.  Look at the number of cancelled flights per day. Is there a pattern?
    Is the proportion of cancelled flights related to the average delay?
    
```{r}
prop_cancelled <- flights %>% 
  group_by( year, month, day ) %>% 
  summarise( prop = mean( is.na(arr_delay) ),
             avg_delay = mean( arr_delay[arr_delay > 0], na.rm = TRUE )) %>% 
  mutate( date = ISOdate(year, month, day) )

ggplot(prop_cancelled) +
  geom_col( aes( x = date, y = prop, fill = weekdays(date))) +
  geom_line(aes( x = date, y=rollmean(prop, 7, fill = NA))) +
  geom_line(aes( x = date, y=rollmean(prop, 30, fill = NA)), color = "red")
```

From the plot it doesn't seem to be a pattern, some days have higher peaks but they are not from the same day (weekday = color).

The moving average on weeks does not highlight anything special, while the monthly one (red) has a couple of spikes around christmas, first week of january (maybe a storm?) and during summer holidays.

```{r}
weekday_cancelled <- flights %>% 
  group_by( year, month, day ) %>% 
  mutate( date = ISOdate(year, month, day),
          weekday = weekdays(date) ) %>% 
  group_by(weekday) %>% 
  summarize( prop = mean( is.na(arr_delay) )) %>% 
  arrange(desc(prop))

weekday_cancelled
```

  The table highlights some differences but we can not discriminate the impact of outliers (e.g. the day where 60% of flights got cancelled for saturdays and fridays)

```{r}
ggplot(prop_cancelled, mapping = aes( x = avg_delay, y = prop)) +
  geom_smooth()
```

For avg_delay greater than 30 min it seems to be a linear correlation. A few outliers are visible in the geom_point plot

1.  Which carrier has the worst delays? Challenge: can you disentangle the
    effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
    `flights %>% group_by(carrier, dest) %>% summarise(n())`)

```{r}
cd_delays <- flights %>% 
  group_by(carrier, dest) %>% 
  summarise( total_delay = sum( dep_delay[dep_delay > 0], na.rm=TRUE) )

ggplot(cd_delays,
       mapping = aes( x = reorder(dest,total_delay), 
                      y = reorder(carrier, total_delay),
                      size = log2(total_delay),
                      color = log2(total_delay))) +
  geom_point() +
  scale_color_distiller(palette="RdYlGn") +
  coord_flip()
```

```{r}
ggplot(cd_delays,
       mapping = aes( x = reorder(dest,total_delay), 
                      y = reorder(carrier, total_delay),
                      fill = log2(total_delay))) +
  geom_tile() +
  scale_fill_distiller(palette="RdYlGn") +
  coord_flip()
```

From this chart it seems that arr_delays greatly depend on the company rather than the departure airport (arrival airport does not seem a plausible hypothesis).

1.  What does the `sort` argument to `count()` do. When might you use it?

  If TRUE, will show the largest groups at the top. Useful not to write arrange(that variable)
    

    
### Exercises

1.  Refer back to the lists of useful mutate and filtering functions. 
    Describe how each operation changes when you combine it with grouping.

1.  Which plane (`tailnum`) has the worst on-time record?

```{r}
delay_chart <- flights %>% 
  select(tailnum, arr_delay) %>% 
  group_by(tailnum) %>%
  summarise( total_flights = n(),
             on_time = ( sum( arr_delay > 0, na.rm = TRUE ) +
                         sum( is.na(arr_delay) ) 
                       ) / length(arr_delay) ) %>%
  filter( !is.na(tailnum)) %>% 
  arrange(on_time)

delay_chart[1:10,] # print worst 10 flights

```

```{r}  
ggplot(delay_chart, mapping = aes(x = on_time)) +
    geom_density() 
```

```{r}
ggplot(delay_chart, mapping = aes(x = total_flights)) +
    geom_density()

```

The problem with the worst flights is that maybe they have flown only few times, so they are maybe unlucky instead of chronically delayed.

```{r}
ggplot(delay_chart, mapping = aes(x = on_time, y = total_flights)) +
  geom_density2d()
```

```{r}
ggplot(filter(delay_chart, total_flights > 20), mapping = aes(x = on_time)) +
    geom_density() 

# cancel low flights noise
delay_chart_filtered <- filter(delay_chart, total_flights > 20)

# fit a beta distribution (probility of probability)
# m <- fitdistr(delay_chart_filtered$on_time, #MASS package
#               dbeta, 
#               start = list(shape1 = 1, shape2 = 10)) 
m <- fitdistrplus::fitdist(delay_chart_filtered$on_time, "beta")

#save parameters
x  <-seq(0,1,length= dim(delay_chart_filtered)[1] )
alpha0 <- m$estimate[1]
beta0  <- m$estimate[2]
db <- dbeta(x, alpha0, beta0)

ggplot(delay_chart_filtered, mapping = aes(x = on_time)) +
    geom_density() +
    geom_line(aes(x,db), color="red")
```

Now er can update our priors

```{r}
delay_chart <- delay_chart %>% 
  mutate( true_on_time = (on_time * total_flights + alpha0) 
                         / (total_flights + alpha0 + beta0 ) ) %>% 
  arrange(true_on_time)
  
delay_chart[1:10,]
```

Under the assumption that all flights are drawn from the same distribution.
This result was obtained by replicating tutorial on http://varianceexplained.org/r/empirical_bayes_baseball/


1.  What time of day should you fly if you want to avoid delays as much
    as possible?
    
```{r}
ggplot(flights) +
  geom_density(aes(dep_time_min))

```

```{r}

hour_delay <- flights %>% 
    group_by(hour) %>% 
    summarise( prob_delay = mean( arr_delay <= 0, na.rm = TRUE),
               total_flights = n() ) %>% 
    arrange(hour)

ggplot(hour_delay) +
  geom_point( aes(hour, prob_delay, color = total_flights > 2000) ) +
  ggtitle("probability( arr_delay <= 0 )")
```

The probability of an on time flights decreases as the hours in the day pass
    
1.  For each destination, compute the total minutes of delay. For each 
    flight, compute the proportion of the total delay for its destination.

```{r}
flights %>% 
  group_by(dest) %>% 
  summarise( tota_delay = sum(arr_delay * (arr_delay > 0), 
                              na.rm = TRUE),
          total_flights = n())
```

```{r}

delay_prop_flights <- flights %>% 
  group_by(dest) %>% 
  mutate( total_delay = sum(arr_delay * (arr_delay > 0), 
                              na.rm = TRUE),
          total_flights = n()) %>% 
  ungroup() %>% 
  group_by(flight) %>% 
  summarise( delay_proportion = (arr_delay > 0) * (arr_delay / total_delay) )

delay_prop_flights 
```

1.  Delays are typically temporally correlated: even once the problem that
    caused the initial delay has been resolved, later flights are delayed 
    to allow earlier flights to leave. Using `lag()`, explore how the delay
    of a flight is related to the delay of the immediately preceding flight.
    
```{r, eval = FALSE}
lag_flights <- flights %>% 
  group_by(origin, year, month, day) %>% 
  summarise( diff = lag(dep_delay) - dep_delay,
             q25 = quantile(diff, 0.25, na.rm = TRUE),
             q50 = quantile(diff, 0.25, na.rm = TRUE),
             q75 = quantile(diff, 0.25, na.rm = TRUE),
             mean = mean(diff, na.rm = TRUE) )

# ok this does not work
```
    
1.  Look at each destination. Can you find flights that are suspiciously
    fast? (i.e. flights that represent a potential data entry error). Compute
    the air time of a flight relative to the shortest flight to that destination.
    Which flights were most delayed in the air?
    
```{r}
flights %>% 
  group_by(dest) %>% 
  mutate( mph =  (distance*60) / air_time ) %>% 
  summarise( max_mph = max(mph, na.rm = TRUE) ) %>% 
  arrange( desc(max_mph) )
```

All velocities seem to be plausible.

Regarding the flights that got delayed the most while they were flying, we must consider
- a non delayed flights takes air_time to travel between two cities
- a dep_delayed flights takes air_time to travel and should arrive at air_time + dep_delay if no further delay is accumulated
- in other words, arr_delay = air_time + dep_delay + mid_air_delay
- or mid_air_delay = arr_delay - (air_time + dep_delay) 

To see which were delayed the most...

```{r}
flights %>% 
  mutate( mid_air_delay = arr_delay - (air_time + dep_delay) ) %>% 
  group_by(flight) %>% 
  summarise( max_mid_air_delay = max(mid_air_delay)) %>% 
  arrange( desc(max_mid_air_delay))
```
    
1.  Find all destinations that are flown by at least two carriers. Use that
    information to rank the carriers.
    
```{r, eval = FALSE}
flights %>% 
  group_by(dest) %>% 
  mutate( more_than_1 = n_distinct(carrier) >= 2 ) %>% 
  ungroup() %>% 
  group_by(carrier, dest) # %>%  don't understand
```

1.  For each plane, count the number of flights before the first delay 
    of greater than 1 hour.
    
```{r}
flights %>% 
  select(tailnum, time_hour, dep_delay) %>% 
  group_by(tailnum) %>% 
  filter( dep_delay > 60 ) %>% 
  filter( rank(time_hour) == 1)
```
